---
title: "Interactive Dendrogram"
format: 
  html:
    page-layout: full
    include-in-header:
      text: |
        <style>
          /* Remove max-width limits from all potential containers */
          .container-fluid, .main-container, .page-columns {
            max-width: 100% !important;
            padding-left: 0 !important;
            padding-right: 0 !important;
          }
        </style>
server: shiny
date: now
date-format: "YYYY-MM-DD HH:mm"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| context: setup

library(shiny)
library(plotly)
library(ggdendro)
library(lsa)
library(mSigPlot)
library(gridExtra)
library(ggplot2)

source("dendrogram_helpers.R")

# Pre-compute both datasets
datasets <- list()
dendrograms <- list()

for (ds in c("Koh89", "Koh476")) {
  datasets[[ds]] <- load_cap9_signatures(ds)
  dendrograms[[ds]] <- compute_dendrogram(datasets[[ds]]$combined)
}
```

```{r}
#| label: wide-layout
tags$style(".container-fluid { max-width: 100% !important; padding: 0 15px; }")
```

This interactive dendrogram lets you explore hierarchical clustering of CAP9 signature extractions.
Zoom into regions of interest and click leaves or use rectangle or lasso select to select leaves to compare signature profiles.

## Controls

```{r}
#| label: ui-controls
#| panel: input

selectInput(
  "dataset",
  "Dataset:",
  choices = c("Koh89" = "Koh89", "Koh476" = "Koh476"),
  selected = "Koh89"
)

uiOutput("selection_display")
```

```{r}
#| label: selection-display-server
#| context: server

# Reactive: selected signatures (max 10)
selected_sigs <- reactiveVal(character(0))

# Clear selection on dataset switch
observeEvent(input$dataset, {
  selected_sigs(character(0))
})

# Render selection badges
output$selection_display <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(tags$p(
      tags$em("Click leaves in the dendrogram to select signatures (max 10).")
    ))
  }

  ds <- input$dataset
  sv <- datasets[[ds]]$source_vec

  badges <- lapply(sigs, function(s) {
    col <- cap9_colors[sv[s]]
    tags$span(
      s,
      style = paste0(
        "display: inline-block; margin: 2px 4px; padding: 3px 8px; ",
        "border-radius: 4px; color: white; font-size: 0.85em; ",
        "background-color: ",
        col,
        ";"
      )
    )
  })

  tagList(
    tags$p(tags$strong(paste0("Selected (", length(sigs), "/10): ")), badges),
    actionButton("clear_selection", "Clear All", style = "margin-top: 4px;")
  )
})

observeEvent(input$clear_selection, {
  selected_sigs(character(0))
})
```


::: {.column-screen}

```{r}
#| label: dendrogram-plot

plotlyOutput("dendrogram", height = "500px", width = "100vw")
```

:::

```{r}
#| label: dendrogram-server
#| context: server

# Render dendrogram only on dataset change (not on selection change,
# which would clear the click event and create a feedback loop)
output$dendrogram <- renderPlotly({
  ds <- input$dataset
  build_plotly_dendrogram(
    dendrograms[[ds]]$dend_data,
    datasets[[ds]]$source_vec,
    colors = cap9_colors
  )
})

# Handle click to toggle signature selection
observeEvent(event_data("plotly_click", source = "dendrogram"), {
  click <- event_data("plotly_click", source = "dendrogram")
  if (!is.null(click)) {
    sig_name <- click$customdata
    if (is.list(sig_name)) {
      sig_name <- sig_name[[1]]
    }
    if (!is.null(sig_name) && nzchar(sig_name)) {
      current <- selected_sigs()
      if (sig_name %in% current) {
        selected_sigs(setdiff(current, sig_name))
      } else if (length(current) < 10) {
        selected_sigs(c(current, sig_name))
      }
    }
  }
})
```

## Signature Profiles

```{r}
#| label: sig-profiles

uiOutput("sig_plots_ui")
```

```{r}
#| label: sig-profiles-server
#| context: server

output$sig_plots_ui <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(tags$p(tags$em("Select signatures above to view their profiles.")))
  }
  height_per <- if (input$dataset == "Koh89") 300 else 400
  total_height <- paste0(length(sigs) * height_per, "px")
  plotOutput("sig_plots", height = total_height, width = "100%")
})

sig_plots_height <- reactive({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(300)
  }
  height_per <- if (input$dataset == "Koh89") 300 else 400
  length(sigs) * height_per
})

output$sig_plots <- renderPlot(
  {
    sigs <- selected_sigs()
    req(length(sigs) > 0)

    ds <- input$dataset
    combined <- datasets[[ds]]$combined
    plot_fn <- if (ds == "Koh89") mSigPlot::plot_89 else mSigPlot::plot_476

    plots <- lapply(sigs, function(s) {
      sig_data <- combined[, s, drop = FALSE]
      plot_fn(sig_data, plot_title = s)
    })

    gridExtra::grid.arrange(grobs = plots, ncol = 1)
  },
  res = 100,
  width = function() {
    if (input$dataset == "Koh476") 1400 else 900
  },
  height = function() {
    sig_plots_height()
  }
)
```

## Cosine Similarity

```{r}
#| label: cosine-heatmap

plotlyOutput("cosine_heatmap", height = "500px")
```

```{r}
#| label: cosine-heatmap-server
#| context: server

output$cosine_heatmap <- renderPlotly({
  sigs <- selected_sigs()
  if (length(sigs) < 2) {
    return(
      plot_ly(type = "scatter", mode = "markers") |>
        layout(
          title = "Select at least 2 signatures to see cosine similarity.",
          xaxis = list(visible = FALSE),
          yaxis = list(visible = FALSE)
        )
    )
  }

  ds <- input$dataset
  cos_mat <- dendrograms[[ds]]$cosine_sim[sigs, sigs]

  # Format for display
  text_mat <- matrix(
    sprintf("%.3f", cos_mat),
    nrow = nrow(cos_mat)
  )

  plot_ly(
    x = sigs,
    y = sigs,
    z = cos_mat,
    text = text_mat,
    type = "heatmap",
    colorscale = list(c(0, "#FFFFFF"), c(1, "#FF0000")),
    zmin = 0,
    zmax = 1,
    hovertemplate = "%{y} vs %{x}<br>Cosine: %{text}<extra></extra>"
  ) |>
    layout(
      title = "Pairwise Cosine Similarity",
      xaxis = list(title = "", tickangle = -45),
      yaxis = list(title = "", autorange = "reversed"),
      margin = list(l = 120, b = 120)
    )
})
```

## About

- **Source colors**: Liu (red), PCAWG (purple), Hartwig cap9 (green), Hartwig nocap (blue)
- **Click** a leaf marker to select/deselect it (shown in badges above).
- **Zoom** into crowded regions using plotly's built-in zoom (drag to select area).
- Signature profiles are rendered with mSigPlot (`plot_89` or `plot_476`).
- Max 10 signatures can be selected at once.
