---
title: "Interactive Signature Assignments"
format: html
server: shiny
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| context: setup

library(shiny)
library(plotly)
library(jsonlite)
library(mSigPlot)
library(gridExtra)
library(ggplot2)
library(ggrepel)

# Load helper functions
source("hamburger_helpers.R")

# Data directory
data_dir <- "../Manuscript_data"

# Load assignment matrices
assignments_83_89 <- read.delim(
  file.path(
    data_dir,
    "assignment_from_172_type",
    "Liu_et_al_83_plus_89_as_83_type_signature_assignments.tsv"
  ),
  row.names = 1,
  check.names = FALSE
)

assignments_648 <- read.delim(
  file.path(data_dir, "Liu_et_al_648_type_signature_assignments.tsv"),
  row.names = 1,
  check.names = FALSE
)

# WARNING: This assignment file is out of date.
# The current 89-type assignments are in
# Manuscript_data/Mo_CAP9_analysis/finalized_cap9/liu_et_al_89_assignment.tsv
# and can be read via read_89_type_assignments() from vhelpers.R.
assignments_89 <- read.delim(
  file.path(
    data_dir,
    "assignment_from_172_type",
    "Liu_et_al_83_plus_89_as_89_type_signature_assignments.tsv"
  ),
  row.names = 1,
  check.names = FALSE
)

assignments_476 <- read.delim(
  file.path(data_dir, "Liu_et_al_476_type_signature_assignments.tsv"),
  row.names = 1,
  check.names = FALSE
)

# Load spectra files
spectra_83 <- read.delim(
  file.path(data_dir, "Liu_et_al_83_type_spectra.tsv"),
  row.names = 1,
  check.names = FALSE
)

spectra_89 <- read.delim(
  file.path(data_dir, "Liu_et_al_89_type_spectra.tsv"),
  row.names = 1,
  check.names = FALSE
)

spectra_476 <- read.delim(
  file.path(data_dir, "Liu_et_al_476_type_spectra.tsv"),
  row.names = 1,
  check.names = FALSE
)

# Load signature files
sigs_83 <- read.delim(
  file.path(data_dir, "Liu_et_al_final_83_type_signatures.tsv"),
  row.names = 1,
  check.names = FALSE
)

sigs_89 <- read.delim(
  file.path(data_dir, "Liu_et_al_final_89_type_signatures.tsv"),
  row.names = 1,
  check.names = FALSE
)

sigs_476 <- read.delim(
  file.path(data_dir, "Liu_et_al_final_476_type_signatures.tsv"),
  row.names = 1,
  check.names = FALSE
)

# Load signature mapping (89-type to 83-type correspondence)
sig_mapping <- read.delim(
  file.path(data_dir, "89type_to_83type_connection.tsv"),
  check.names = FALSE
)

# Load sample info for MSI status coloring
sample_info <- read.delim(
  file.path(data_dir, "sample_info.tsv"),
  check.names = FALSE
)

# Get signature lists
signatures_83_89 <- rownames(assignments_83_89)
signatures_89 <- rownames(assignments_89)
signatures_476 <- rownames(assignments_476)
signatures_648 <- rownames(assignments_648)
```

This interactive visualization shows the distribution of signature mutations across cancer types ("hamburger plots"). Click on any point to view that tumor's mutational spectrum in ID83, ID89, and ID476 classifications using mSigPlot.

## Select Assignment Type and Signature

```{r}
#| label: ui-controls
#| panel: input

selectInput(
  "assignment_type",
  "Assignment Type:",
  choices = c(
    "83+89 as 83-type (33 signatures)" = "83_89",
    "83+89 as 89-type (48 signatures)" = "89",
    "476-type (44 signatures)" = "476",
    "648-type (44 signatures)" = "648"
  ),
  selected = "83_89"
)

uiOutput("signature_selector")
```

```{r}
#| label: signature-selector-server
#| context: server

output$signature_selector <- renderUI({
  sigs <- switch(
    input$assignment_type,
    "83_89" = signatures_83_89,
    "89" = signatures_89,
    "476" = signatures_476,
    "648" = signatures_648,
    signatures_83_89  # default
  )
  selectInput("signature", "Signature:", choices = sigs)
})
```

## Hamburger Plot

Click on any point to select a tumor and view its spectrum below.

```{r}
#| label: hamburger-plot

plotlyOutput("hamburger_plot", height = "500px")
```

```{r}
#| label: hamburger-plot-server
#| context: server

# Helper function to map signature names between assignment types
# 83_89 uses C_ID* names, while 89/476/648 use InsDel* names
map_signature_name <- function(sig_name, from_type, to_type, mapping) {

  if (from_type == to_type) {
    return(sig_name)
  }

  # Determine if we need to convert between naming conventions
  from_is_83 <- from_type == "83_89"
  to_is_83 <- to_type == "83_89"

  if (from_is_83 && !to_is_83) {
    # Map from 83-type (C_ID*) to 89-type (InsDel*)
    matches <- mapping$InDel89[mapping$InDel83 == sig_name]
    if (length(matches) > 0) {
      # Return the one with lowest lexicographic order
      return(sort(matches)[1])
    }
  } else if (!from_is_83 && to_is_83) {
    # Map from 89-type (InsDel*) to 83-type (C_ID*)
    matches <- mapping$InDel83[mapping$InDel89 == sig_name]
    if (length(matches) > 0) {
      return(matches[1])
    }
  }
  # If both use same naming convention (89/476/648), return as-is

  # Return original if no mapping found
  return(sig_name)
}

# Reactive to get the mapped signature name for current assignment type
mapped_signature <- reactive({
  req(input$signature, input$assignment_type)

  sig_name <- input$signature

  # Determine the source type based on whether the signature exists in each matrix
  if (sig_name %in% rownames(assignments_83_89)) {
    source_type <- "83_89"
  } else if (sig_name %in% rownames(assignments_89)) {
    source_type <- "89"
  } else if (sig_name %in% rownames(assignments_476)) {
    source_type <- "476"
  } else if (sig_name %in% rownames(assignments_648)) {
    source_type <- "648"
  } else {
    return(sig_name)
  }

  map_signature_name(sig_name, source_type, input$assignment_type, sig_mapping)
})

# Reactive for current hamburger data
hamburger_data <- reactive({
  sig_name <- mapped_signature()
  req(sig_name)

  assignment_matrix <- switch(
    input$assignment_type,
    "83_89" = assignments_83_89,
    "89" = assignments_89,
    "476" = assignments_476,
    "648" = assignments_648,
    assignments_83_89  # default
  )

  # Check if signature exists in the matrix
  if (!sig_name %in% rownames(assignment_matrix)) {
    return(NULL)
  }

  get_hamburger_data(sig_name, assignment_matrix, min_samples = 3)
})

# Render hamburger plot
output$hamburger_plot <- renderPlotly({
  df <- hamburger_data()

  if (is.null(df) || nrow(df) == 0) {
    return(
      plotly_empty() |> layout(title = "No data available for this signature")
    )
  }

  type_order <- attr(df, "type_order")
  medians <- attr(df, "medians")
  n_types <- length(type_order)

  bg_colors <- c("#EDF8B1", "#2C7FB8")

  # Create alternating background shapes
  bg_shapes <- lapply(seq_len(n_types), function(i) {
    list(
      type = "rect",
      x0 = i - 0.5,
      x1 = i + 0.5,
      y0 = 0,
      y1 = 1,
      xref = "x",
      yref = "paper",
      fillcolor = bg_colors[(i - 1) %% 2 + 1],
      opacity = 0.2,
      line = list(width = 0),
      layer = "below"
    )
  })

  # Create median line shapes
  median_shapes <- lapply(seq_len(nrow(medians)), function(i) {
    x_pos <- which(type_order == medians$cancer_type[i])
    list(
      type = "line",
      x0 = x_pos - 0.4,
      x1 = x_pos + 0.4,
      y0 = medians$median_mutations[i],
      y1 = medians$median_mutations[i],
      xref = "x",
      yref = "y",
      line = list(color = "red", width = 2, dash = "dash"),
      layer = "above"
    )
  })

  all_shapes <- c(bg_shapes, median_shapes)

  # Extract patient ID from sample_id (remove "CancerType::" prefix)
  df$patient_id <- sub("^.*::", "", df$sample_id)

  # Look up MSI status from sample_info
  df$msi_status <- sample_info$`MSIseq_MSI-H`[match(df$patient_id, sample_info$Patient)]

  # Set dot color: red for MSI-H (TRUE), black for others
 df$dot_color <- ifelse(!is.na(df$msi_status) & df$msi_status == TRUE, "red", "black")

  # Create hover text
  df$hover_text <- paste0(
    "<b>Sample:</b> ",
    df$sample_id,
    "<br>",
    "<b>Cancer Type:</b> ",
    df$cancer_type,
    "<br>",
    "<b>Mutations:</b> ",
    round(df$mutations, 1),
    "<br>",
    "<b>MSI-H:</b> ",
    ifelse(!is.na(df$msi_status) & df$msi_status == TRUE, "Yes", "No")
  )

  # Create title showing mapped signature name
  sig_display <- mapped_signature()
  if (sig_display != input$signature) {
    plot_title <- paste0(sig_display, " (from ", input$signature, ") mutations")
  } else {
    plot_title <- paste0(sig_display, " mutations")
  }

  plot_ly(
    data = df,
    x = ~x_plot,
    y = ~mutations,
    type = "scatter",
    mode = "markers",
    marker = list(size = 6, color = ~dot_color, opacity = 0.6),
    text = ~hover_text,
    hoverinfo = "text",
    customdata = ~sample_id,
    source = "hamburger"
  ) |>
    layout(
      title = list(text = plot_title, x = 0.5),
      xaxis = list(
        title = "",
        tickmode = "array",
        tickvals = seq_len(n_types),
        ticktext = as.character(type_order),
        tickangle = 60,
        side = "top",
        showgrid = FALSE
      ),
      yaxis = list(
        title = "Number of Mutations",
        type = "log",
        showgrid = TRUE
      ),
      shapes = all_shapes,
      hovermode = "closest",
      margin = list(t = 120, b = 50, l = 60, r = 20)
    ) |>
    config(displayModeBar = TRUE, displaylogo = FALSE) |>
    event_register("plotly_click")
})

# Track selected sample from click
selected_sample <- reactiveVal(NULL)

# Track whether to show decomposition view
show_decomposition <- reactiveVal(FALSE)

observeEvent(event_data("plotly_click", source = "hamburger"), {
  click_data <- event_data("plotly_click", source = "hamburger")
  if (!is.null(click_data)) {
    selected_sample(click_data$customdata[[1]])
    show_decomposition(FALSE) # Reset to spectrum view when new sample selected
  }
})

# Handle click on sample ID to show decomposition
observeEvent(input$sample_id_link, {
  show_decomposition(TRUE)
})

# Handle click on back button to return to spectrum view
observeEvent(input$back_to_spectrum, {
  show_decomposition(FALSE)
})
```

## Selected Tumor Spectrum

```{r}
#| label: selected-info

uiOutput("selected_info")
```

```{r}
#| label: selected-info-server
#| context: server

output$selected_info <- renderUI({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    tags$p(tags$strong(
      "Select a point from the hamburger plot above to view spectrum."
    ))
  } else if (show_decomposition()) {
    tags$div(
      tags$p(
        tags$strong("Selected Tumor: "),
        sample_id,
        " - Showing signature decomposition"
      ),
      actionButton(
        "back_to_spectrum",
        "Back to Spectrum View",
        style = "margin-top: 5px;"
      )
    )
  } else {
    tags$p(
      tags$strong("Selected Tumor: "),
      actionLink(
        "sample_id_link",
        sample_id,
        style = "color: blue; text-decoration: underline; cursor: pointer;"
      ),
      tags$span(
        " (click to see signature decomposition)",
        style = "color: gray; font-size: 0.9em;"
      )
    )
  }
})
```

::: {.panel-tabset}

### 83-type

```{r}
#| label: spectrum-83

uiOutput("spectrum_83_ui")
```

```{r}
#| label: spectrum-83-server
#| context: server

# Dynamic UI for 83-type plot with adjustable height
output$spectrum_83_ui <- renderUI({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    return(tags$p("Select a sample to view spectrum."))
  }

  # Calculate height based on number of non-zero signatures in decomposition mode
  if (show_decomposition() && sample_id %in% colnames(assignments_83_89)) {
    sample_assignments <- assignments_83_89[, sample_id]
    n_sigs <- sum(sample_assignments > 0)
    # 400px donut + 300px spectrum + 300px reconstructed + 300px per signature
    height <- paste0(400 + 300 + 300 + n_sigs * 300, "px")
  } else {
    height <- "500px"
  }

  plotOutput("spectrum_83", height = height)
})

output$spectrum_83 <- renderPlot(
  {
    sample_id <- selected_sample()
    sig_name <- input$signature
    req(sample_id, sig_name)

    if (!sample_id %in% colnames(spectra_83)) {
      plot.new()
      text(0.5, 0.5, paste("No ID83 data for", sample_id))
      return()
    }

    # Check if we should show decomposition view
    if (show_decomposition()) {
      plot_decomposition(
        sample_id = sample_id,
        assignments = assignments_83_89,
        signatures = sigs_83,
        spectra = spectra_83,
        plot_fn = mSigPlot::plot_83,
        title_prefix = "ID83"
      )
      return()
    }

    # Normal spectrum view
    # Get the 83-type signature name
    if (input$assignment_type == "83_89") {
      sig_83_name <- sig_name # Already 83-type
    } else {
      # Map from 89-type to 83-type
      idx <- match(sig_name, sig_mapping$InDel89)
      sig_83_name <- if (!is.na(idx)) sig_mapping$InDel83[idx] else NULL
    }

    spectrum <- spectra_83[, sample_id, drop = FALSE]
    p1 <- mSigPlot::plot_83(
      spectrum,
      plot_title = paste("Spectrum:", sample_id)
    )

    if (!is.null(sig_83_name) && sig_83_name %in% colnames(sigs_83)) {
      sig_data <- sigs_83[, sig_83_name, drop = FALSE]
      p2 <- mSigPlot::plot_83(
        sig_data,
        plot_title = paste("Signature:", sig_83_name)
      )
      gridExtra::grid.arrange(p1, p2, ncol = 1)
    } else {
      p1
    }
  },
  res = 100
)
```

### 89-type

```{r}
#| label: spectrum-89

uiOutput("spectrum_89_ui")
```

```{r}
#| label: spectrum-89-server
#| context: server

# Dynamic UI for 89-type plot with adjustable height
output$spectrum_89_ui <- renderUI({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    return(tags$p("Select a sample to view spectrum."))
  }

  # Calculate height based on number of non-zero signatures in decomposition mode
  if (show_decomposition() && sample_id %in% colnames(assignments_89)) {
    sample_assignments <- assignments_89[, sample_id]
    n_sigs <- sum(sample_assignments > 0)
    # 400px donut + 300px spectrum + 300px reconstructed + 300px per signature
    height <- paste0(400 + 300 + 300 + n_sigs * 300, "px")
  } else {
    height <- "500px"
  }

  plotOutput("spectrum_89", height = height)
})

output$spectrum_89 <- renderPlot(
  {
    sample_id <- selected_sample()
    sig_name <- input$signature
    req(sample_id, sig_name)

    if (!sample_id %in% colnames(spectra_89)) {
      plot.new()
      text(0.5, 0.5, paste("No ID89 data for", sample_id))
      return()
    }

    # Check if we should show decomposition view
    if (show_decomposition()) {
      plot_decomposition(
        sample_id = sample_id,
        assignments = assignments_89,
        signatures = sigs_89,
        spectra = spectra_89,
        plot_fn = mSigPlot::plot_89,
        title_prefix = "ID89"
      )
      return()
    }

    # Normal spectrum view
    # Get the 89-type signature name
    if (input$assignment_type == "648") {
      sig_89_name <- sig_name # Already 89-type
    } else {
      # Map from 83-type to 89-type (find first matching 89-type)
      idx <- which(sig_mapping$InDel83 == sig_name)
      sig_89_name <- if (length(idx) > 0) sig_mapping$InDel89[idx[1]] else NULL
    }

    spectrum <- spectra_89[, sample_id, drop = FALSE]
    p1 <- mSigPlot::plot_89(
      spectrum,
      plot_title = paste("Spectrum:", sample_id)
    )

    if (!is.null(sig_89_name) && sig_89_name %in% colnames(sigs_89)) {
      sig_data <- sigs_89[, sig_89_name, drop = FALSE]
      p2 <- mSigPlot::plot_89(
        sig_data,
        plot_title = paste("Signature:", sig_89_name)
      )
      gridExtra::grid.arrange(p1, p2, ncol = 1)
    } else {
      p1
    }
  },
  res = 100
)
```

### 476-type

```{r}
#| label: spectrum-476

uiOutput("spectrum_476_ui")
```

```{r}
#| label: spectrum-476-server
#| context: server

# Dynamic UI for 476-type plot with adjustable height
output$spectrum_476_ui <- renderUI({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    return(tags$p("Select a sample to view spectrum."))
  }

  # Calculate height based on number of non-zero signatures in decomposition mode
  if (show_decomposition() && sample_id %in% colnames(assignments_476)) {
    sample_assignments <- assignments_476[, sample_id]
    n_sigs <- sum(sample_assignments > 0)
    # 500px donut + 400px spectrum + 400px reconstructed + 400px per signature (476-type plots need more height)
    height <- paste0(500 + 400 + 400 + n_sigs * 400, "px")
  } else {
    height <- "600px"
  }

  plotOutput("spectrum_476", height = height, width = "100%")
})

# Reactive to calculate 476 plot height for renderPlot
spectrum_476_height <- reactive({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    return(600)
  }

  if (show_decomposition() && sample_id %in% colnames(assignments_476)) {
    sample_assignments <- assignments_476[, sample_id]
    n_sigs <- sum(sample_assignments > 0)
    # 500px donut + 400px spectrum + 400px reconstructed + 400px per signature
    return(500 + 400 + 400 + n_sigs * 400)
  }
  600
})

output$spectrum_476 <- renderPlot(
  {
    sample_id <- selected_sample()
    sig_name <- input$signature
    req(sample_id, sig_name)

    if (!sample_id %in% colnames(spectra_476)) {
      plot.new()
      text(0.5, 0.5, paste("No ID476 data for", sample_id))
      return()
    }

    # Check if we should show decomposition view
    if (show_decomposition()) {
      plot_decomposition(
        sample_id = sample_id,
        assignments = assignments_476,
        signatures = sigs_476,
        spectra = spectra_476,
        plot_fn = mSigPlot::plot_476,
        title_prefix = "ID476"
      )
      return()
    }

    # Normal spectrum view
    # Get the 476-type signature name (uses same names as 89-type)
    if (input$assignment_type == "648") {
      # Already 89-type name, use directly
      sig_476_name <- sig_name
    } else {
      # Map from 83-type to 89-type (find first matching)
      idx <- which(sig_mapping$InDel83 == sig_name)
      sig_476_name <- if (length(idx) > 0) sig_mapping$InDel89[idx[1]] else NULL
    }

    spectrum <- spectra_476[, sample_id, drop = FALSE]
    p1 <- mSigPlot::plot_476(
      spectrum,
      plot_title = paste("Spectrum:", sample_id)
    )

    if (!is.null(sig_476_name) && sig_476_name %in% colnames(sigs_476)) {
      sig_data <- sigs_476[, sig_476_name, drop = FALSE]
      p2 <- mSigPlot::plot_476(
        sig_data,
        plot_title = paste("Signature:", sig_476_name)
      )
      gridExtra::grid.arrange(p1, p2, ncol = 1)
    } else {
      p1
    }
  },
  res = 100,
  width = 1400,
  height = function() {
    spectrum_476_height()
  }
)
```

### 648 assignments / 476 plots

```{r}
#| label: spectrum-648-as-476

uiOutput("spectrum_648_476_ui")
```

```{r}
#| label: spectrum-648-as-476-server
#| context: server

# Dynamic UI for 648-assignments with 476-type plots
output$spectrum_648_476_ui <- renderUI({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    return(tags$p("Select a sample to view spectrum."))
  }

  # Calculate height based on number of non-zero signatures in decomposition mode
  if (show_decomposition() && sample_id %in% colnames(assignments_648)) {
    sample_assignments <- assignments_648[, sample_id]
    n_sigs <- sum(sample_assignments > 0)
    # 500px donut + 400px spectrum + 400px reconstructed + 400px per signature
    height <- paste0(500 + 400 + 400 + n_sigs * 400, "px")
  } else {
    height <- "600px"
  }

  plotOutput("spectrum_648_476", height = height, width = "100%")
})

# Reactive to calculate plot height for renderPlot
spectrum_648_476_height <- reactive({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    return(600)
  }

  if (show_decomposition() && sample_id %in% colnames(assignments_648)) {
    sample_assignments <- assignments_648[, sample_id]
    n_sigs <- sum(sample_assignments > 0)
    return(500 + 400 + 400 + n_sigs * 400)
  }
  600
})

output$spectrum_648_476 <- renderPlot(
  {
    sample_id <- selected_sample()
    sig_name <- input$signature
    req(sample_id, sig_name)

    if (!sample_id %in% colnames(spectra_476)) {
      plot.new()
      text(0.5, 0.5, paste("No ID476 data for", sample_id))
      return()
    }

    # Check if we should show decomposition view
    if (show_decomposition()) {
      plot_decomposition(
        sample_id = sample_id,
        assignments = assignments_648,
        signatures = sigs_476,
        spectra = spectra_476,
        plot_fn = mSigPlot::plot_476,
        title_prefix = "648 as 476"
      )
      return()
    }

    # Normal spectrum view
    # Get the signature name (648-type uses same names as 476-type)
    if (input$assignment_type == "648") {
      sig_476_name <- sig_name
    } else {
      # Map from 83-type to 89-type (find first matching)
      idx <- which(sig_mapping$InDel83 == sig_name)
      sig_476_name <- if (length(idx) > 0) sig_mapping$InDel89[idx[1]] else NULL
    }

    spectrum <- spectra_476[, sample_id, drop = FALSE]
    p1 <- mSigPlot::plot_476(
      spectrum,
      plot_title = paste("Spectrum:", sample_id)
    )

    if (!is.null(sig_476_name) && sig_476_name %in% colnames(sigs_476)) {
      sig_data <- sigs_476[, sig_476_name, drop = FALSE]
      p2 <- mSigPlot::plot_476(
        sig_data,
        plot_title = paste("Signature:", sig_476_name)
      )
      gridExtra::grid.arrange(p1, p2, ncol = 1)
    } else {
      p1
    }
  },
  res = 100,
  width = 1400,
  height = function() {
    spectrum_648_476_height()
  }
)
```

:::

## About This Visualization

This page displays **hamburger plots** (also known as snake plots) showing how mutations attributed to each signature are distributed across cancer types.

### How to Use

1. **Select an assignment type**: Choose between the 83+89 as 83-type assignments (33 signatures) or the 648-type assignments (44 signatures).

2. **Select a signature**: Pick a signature from the dropdown to view its hamburger plot.

3. **Click on a point**: Each point represents one tumor. Click on a point to view that tumor's full mutational spectrum rendered with mSigPlot in ID83, ID89, and ID476 classifications.

4. **View signature decomposition**: Click on the sample ID (shown in blue) to see how that tumor's spectrum is decomposed into constituent signatures. This shows a bar plot of mutation counts assigned to each signature, followed by plots of the individual signatures.

### Reading the Hamburger Plot

- **X-axis**: Cancer types, ordered by median mutation count (ascending)
- **Y-axis**: Number of mutations attributed to this signature (log scale)
- **Points**: Individual tumors, spread horizontally within each cancer type column
- **Red dashed lines**: Median mutation count for each cancer type
- **Alternating backgrounds**: Help distinguish adjacent cancer type columns
