---
title: "Interactive Dendrogram"
format: 
  html:
    page-layout: full
    include-in-header:
      text: |
        <style>
          /* This allows the .column-screen class to work in Shiny mode */
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
          /* Ensure the plot container itself isn't padded */
          .column-screen {
            width: 100% !important;
            margin: 0 !important;
          }
        </style>
server: shiny
date: now
date-format: "YYYY-MM-DD HH:mm"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| context: setup

library(shiny)
library(plotly)
library(ggdendro)
library(lsa)
library(mSigPlot)
library(gridExtra)
library(ggplot2)

source("dendro2_helpers.R")

# Pre-load both full datasets
all_data <- list()
for (ds in c("Koh89", "Koh476", "COSMIC83")) {
  all_data[[ds]] <- load_all_signatures(ds)
}

# Signature names grouped by dataset and source group
sig_names_by_source <- list()
for (ds in c("Koh89", "Koh476", "COSMIC83")) {
  sv <- all_data[[ds]]$source_vec
  sig_names_by_source[[ds]] <- split(names(sv), sv)
}
```


This interactive dendrogram loads **all** CAP9/NoCAP signature extractions and lets you
pick a seed signature from any source group. Only signatures with cosine similarity >= threshold to
the seed are shown in the dendrogram.

## Controls

```{r}
#| label: ui-controls
#| panel: input

selectInput(
  "dataset",
  "Dataset:",
  choices = c("Koh89" = "Koh89", "Koh476" = "Koh476", "COSMIC83" = "COSMIC83"),
  selected = "Koh89"
)

selectInput(
  "seed_source",
  "Seed Source:",
  choices = names(sig_names_by_source[["Koh89"]]),
  selected = "Liu"
)

selectInput(
  "seed_signature",
  "Seed Signature:",
  choices = sig_names_by_source[["Koh89"]][["Liu"]]
)

numericInput(
  "cosine_sim_to_seed",
  "Min Cosine Similarity to Seed:",
  value = 0.8,
  min = 0,
  max = 1,
  step = 0.05
)

uiOutput("source_type_checkboxes")

actionButton("update_dendro", "Update Dendrogram")
actionButton("select_all", "Select All Leaves")

uiOutput("selection_display")
```

```{r}
#| label: server-logic
#| context: server

# Update source group dropdown when dataset changes
observeEvent(input$dataset, {
  sources <- names(sig_names_by_source[[input$dataset]])
  updateSelectInput(session, "seed_source", choices = sources, selected = "Liu")
})

# Update seed signature dropdown when dataset or source group changes
observeEvent(list(input$dataset, input$seed_source), {
  ds <- input$dataset
  src <- input$seed_source
  sigs <- sig_names_by_source[[ds]][[src]]
  updateSelectInput(session, "seed_signature", choices = sigs)
})

# Display labels for source groups
source_display_labels <- c(
  Liu = "Liu",
  C.H = "FMH",
  C.P = "PCAWG",
  C.PH = "PCAWG+FMH",
  N.H = "FMH",
  Koh = "Koh",
  COSMIC = "COSMIC v3.5",
  Liu.m = "Liu (mapped from 476)",
  C.H.m = "FMH (mapped from 476)",
  C.P.m = "PCAWG (mapped from 476)",
  C.PH.m = "PCAWG+FMH (mapped from 476)",
  N.H.m = "FMH (mapped from 476)",
  Sp.C = "Spectra",
  Sp.N = "Spectra",
  `~Sp.C` = "Spectra (from matches in other indel type)",
  `~Sp.N` = "Spectra (from matches in other indel type)"
)

# Render source type checkboxes dynamically based on dataset
output$source_type_checkboxes <- renderUI({
  ds <- input$dataset
  all_sources <- unique(all_data[[ds]]$source_vec)
  default_off <- grepl("\\.m$", all_sources) |
    grepl("^~?Sp\\.", all_sources) |
    all_sources == "N.H"
  selected <- all_sources[!default_off]

  # Build named choices (display label = internal value)
  make_choices <- function(srcs) {
    labels <- ifelse(
      srcs %in% names(source_display_labels),
      source_display_labels[srcs],
      srcs
    )
    setNames(srcs, labels)
  }

  # Define ordering: Capped (C.*) + Liu + Koh first, then Uncapped (N.*)
  capped_order <- c(
    "Liu",
    "Koh",
    "COSMIC",
    "C.H",
    "C.P",
    "C.PH",
    "C.H.m",
    "C.P.m",
    "C.PH.m",
    "Liu.m",
    "Sp.C",
    "~Sp.C"
  )
  uncapped_order <- c("N.H", "N.H.m", "Sp.N", "~Sp.N")

  capped_sources <- intersect(capped_order, all_sources)
  uncapped_sources <- intersect(uncapped_order, all_sources)

  tagList(
    tags$strong("Capped (R ≤ 9):"),
    checkboxGroupInput(
      "source_types_capped",
      label = NULL,
      choices = make_choices(capped_sources),
      selected = intersect(capped_sources, selected),
      inline = TRUE
    ),
    tags$strong("Uncapped:"),
    checkboxGroupInput(
      "source_types_uncapped",
      label = NULL,
      choices = make_choices(uncapped_sources),
      selected = intersect(uncapped_sources, selected),
      inline = TRUE
    )
  )
})

# Reactive: selected signatures (max 10)
selected_sigs <- reactiveVal(character(0))

# Clear selection on dataset switch or dendro update
observeEvent(input$dataset, {
  selected_sigs(character(0))
})

# Filtered + clustered data, recomputed only on button click
filtered_result <- eventReactive(
  input$update_dendro,
  {
    ds <- isolate(input$dataset)
    seed <- isolate(input$seed_signature)
    threshold <- isolate(input$cosine_sim_to_seed)
    active_sources <- c(
      isolate(input$source_types_capped),
      isolate(input$source_types_uncapped)
    )

    combined <- all_data[[ds]]$combined
    source_vec <- all_data[[ds]]$source_vec

    # Filter to active source types (always keep the seed signature)
    # When NULL (initial render before UI exists), use defaults: exclude .m and Sp. groups
    if (is.null(active_sources)) {
      all_sources <- unique(source_vec)
      default_off <- grepl("\\.m$", all_sources) |
        grepl("^~?Sp\\.", all_sources) |
        all_sources == "N.H"
      active_sources <- all_sources[!default_off]
    }
    keep_cols <- names(source_vec)[source_vec %in% active_sources]
    if (!(seed %in% keep_cols)) {
      keep_cols <- c(seed, keep_cols)
    }
    combined <- combined[, keep_cols, drop = FALSE]
    source_vec <- source_vec[keep_cols]

    filtered <- filter_by_seed(combined, seed, threshold)
    filtered_source <- source_vec[colnames(filtered)]

    dendro <- compute_dendrogram(filtered)

    selected_sigs(character(0))

    list(
      combined = filtered,
      source_vec = filtered_source,
      dendro = dendro
    )
  },
  ignoreNULL = FALSE
)

# Render selection badges
output$selection_display <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(tags$p(
      tags$em(
        "Click leaves in the dendrogram or use the rectangle or lasso select tool to select signatures for plotting below."
      )
    ))
  }

  res <- filtered_result()
  sv <- res$source_vec

  badges <- lapply(sigs, function(s) {
    col <- dendro2_colors[sv[s]]
    tags$span(
      s,
      style = paste0(
        "display: inline-block; margin: 2px 4px; padding: 3px 8px; ",
        "border-radius: 4px; color: white; font-size: 0.85em; ",
        "background-color: ",
        col,
        ";"
      )
    )
  })

  tagList(
    tags$p(tags$strong(paste0("Selected (", length(sigs), "): ")), badges),
    actionButton("clear_selection", "Clear All", style = "margin-top: 4px;")
  )
})

observeEvent(input$clear_selection, {
  selected_sigs(character(0))
})

observeEvent(input$select_all, {
  res <- filtered_result()
  selected_sigs(colnames(res$combined))
})
```


::: {.column-screen}

```{r}
#| label: dendrogram-plot

plotlyOutput("dendrogram", height = "500px", width = "100%")
```

:::

```{r}
#| label: dendrogram-server
#| context: server

output$dendrogram <- renderPlotly({
  res <- filtered_result()
  build_plotly_dendrogram(
    res$dendro$dend_data,
    res$source_vec,
    colors = dendro2_colors
  )
})

# Handle click to toggle signature selection
observeEvent(event_data("plotly_click", source = "dendrogram"), {
  click <- event_data("plotly_click", source = "dendrogram")
  if (!is.null(click)) {
    sig_name <- click$customdata
    if (is.list(sig_name)) {
      sig_name <- sig_name[[1]]
    }
    if (!is.null(sig_name) && nzchar(sig_name)) {
      current <- selected_sigs()
      if (sig_name %in% current) {
        selected_sigs(setdiff(current, sig_name))
      } else {
        selected_sigs(c(current, sig_name))
      }
    }
  }
})

# Handle lasso/box selection of leaf markers
observeEvent(event_data("plotly_selected", source = "dendrogram"), {
  sel <- event_data("plotly_selected", source = "dendrogram")
  if (!is.null(sel) && !is.null(sel$customdata)) {
    sig_names <- unlist(sel$customdata)
    sig_names <- sig_names[!is.na(sig_names) & nzchar(sig_names)]
    if (length(sig_names) > 0) {
      selected_sigs(unique(sig_names))
    }
  }
})
```

## Signature Profiles

```{r}
#| label: sig-profiles

uiOutput("sig_plots_ui")
```

```{r}
#| label: sig-profiles-server
#| context: server

output$sig_plots_ui <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(tags$p(tags$em("Select signatures above to view their profiles.")))
  }
  height_per <- if (input$dataset == "Koh476") 400 else 300
  total_height <- paste0(length(sigs) * height_per, "px")
  plotOutput("sig_plots", height = total_height, width = "100%")
})

sig_plots_height <- reactive({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(300)
  }
  height_per <- if (input$dataset == "Koh476") 400 else 300
  length(sigs) * height_per
})

output$sig_plots <- renderPlot(
  {
    sigs <- selected_sigs()
    req(length(sigs) > 0)

    res <- filtered_result()
    leaf_order <- res$dendro$hc$labels[res$dendro$hc$order]
    sigs <- sigs[order(match(sigs, leaf_order))]
    combined <- res$combined
    plot_fn <- if (input$dataset == "Koh89") {
      mSigPlot::plot_89
    } else if (input$dataset == "COSMIC83") {
      mSigPlot::plot_83
    } else {
      mSigPlot::plot_476
    }

    plots <- lapply(sigs, function(s) {
      sig_data <- combined[, s, drop = FALSE]
      plot_fn(sig_data, plot_title = s)
    })

    gridExtra::grid.arrange(grobs = plots, ncol = 1)
  },
  res = 100,
  width = function() {
    if (input$dataset == "Koh476") 1400 else 900 # Koh89 and COSMIC83 both use 900
  },
  height = function() {
    sig_plots_height()
  }
)
```

## Cosine Similarity

```{r}
#| label: cosine-table

uiOutput("cosine_table")
```

```{r}
#| label: cosine-table-server
#| context: server

output$cosine_table <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) < 2) {
    return(tags$p(tags$em(
      "Select at least 2 signatures to see cosine similarity."
    )))
  }

  res <- filtered_result()
  leaf_order <- res$dendro$hc$labels[res$dendro$hc$order]
  sigs <- sigs[order(match(sigs, leaf_order))]
  cos_mat <- res$dendro$cosine_sim[sigs, sigs]
  n <- length(sigs)

  # Build body rows (lower triangle: row i shows columns 1..i-1)
  body_rows <- lapply(2:n, function(i) {
    cells <- list(tags$td(
      style = "font-weight: bold; white-space: nowrap; padding: 4px; font-size: 0.85em;",
      sigs[i]
    ))
    for (j in seq_len(i - 1)) {
      val <- cos_mat[i, j]
      bg <- if (val >= 0.9) {
        "#ffcccc"
      } else if (val >= 0.8) {
        "#ffe0cc"
      } else {
        "#ffffff"
      }
      cells <- c(
        cells,
        list(tags$td(
          style = paste0(
            "text-align: center; padding: 4px; font-size: 0.85em; background-color: ",
            bg,
            ";"
          ),
          sprintf("%.3f", val)
        ))
      )
    }
    do.call(tags$tr, cells)
  })

  # Build footer row with vertical column labels (skip last sig — it's only a row)
  footer_cells <- list(tags$td(""))
  for (j in seq_len(n - 1)) {
    footer_cells <- c(
      footer_cells,
      list(tags$td(
        style = "writing-mode: vertical-rl; transform: rotate(180deg); white-space: nowrap; padding: 4px; font-size: 0.85em; font-weight: bold;",
        sigs[j]
      ))
    )
  }
  footer_row <- do.call(tags$tr, footer_cells)

  tags$table(
    style = "border-collapse: collapse; border: 1px solid #ddd;",
    do.call(tags$tbody, body_rows),
    tags$tfoot(footer_row)
  )
})
```

## About

- **Source colors**: Liu (red), CAP9 FMH (green), CAP9 PCAWG (purple), CAP9 PCAWG+FMH (orange), NoCAP FMH (blue), Koh (brown); "(mapped)" groups are 476→89 mapped versions (lighter shades); Spectra CAP9 = CAP9 best-match spectra (teal), Spectra NoCAP = NoCAP best-match spectra (crimson)
- **Workflow**: Select a dataset, pick a seed source and signature, set cosine similarity threshold, click "Update Dendrogram"
- **Click** a leaf marker to select/deselect it (shown in badges above)
- **Zoom** into crowded regions using plotly's built-in zoom
- Signature profiles are rendered with mSigPlot (`plot_89` or `plot_476`)
- No limit on the number of signatures that can be selected
